---
title: "Analysis of Old Faithful Data"
author: "SDW Yates"
date: "2025-01-26"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: united
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Old Faithful Geyser Analysis

## Load Necessary Libraries
```{r libraries, include=TRUE, warning=FALSE}
library(ggplot2)
library(MASS)
library(stats)
```

## Load the Old Faithful Dataset
```r
data("faithful")

# Preview the dataset
head(faithful)
```
Output:
```
  eruptions waiting
1      3.60      79
2      1.80      54
3      3.33      74
4      2.28      62
5      4.53      85
6      2.88      55
```

## Water-Like Appearance for Eruption Times
```{r eruption_histogram}
eruption_hist <- ggplot(faithful, aes(x = eruptions)) +
  geom_histogram(aes(y = after_stat(density)), bins = 20, fill = "lightblue", color = "blue") +
  geom_density(color = "navy", linewidth = 1.2) +
  labs(title = "Eruption Times: Water-Themed Histogram",
       x = "Eruption Time (mins)", y = "Density") +
  theme_minimal()

eruption_hist
```
The density histogram of eruption times reveals a bimodal distribution, with peaks around 2 and 4.5 minutes. This suggests two primary eruption patterns: short and long durations.


## Explosion-Like Appearance for Waiting Times
```{r waiting_histogram}
waiting_hist <- ggplot(faithful, aes(x = waiting)) +
  geom_histogram(aes(y = after_stat(density)), bins = 20, fill = "orange", color = "red") +
  geom_density(color = "darkred", linewidth = 1.2) +
  labs(title = "Waiting Times: Explosion-Themed Histogram",
       x = "Waiting Time (mins)", y = "Density") +
  theme_minimal()

waiting_hist
```
The density histogram of waiting times shows a unimodal distribution, with a peak near 80 minutes. Most waiting times cluster around this value, reflecting consistent intervals between eruptions.


## Combined Density Histogram for Eruption and Waiting Times
```{r combined_histograms, fig.width=11, fig.height=6}
combined_density <- ggplot() +
  geom_density(data = faithful, aes(x = eruptions, y = after_stat(density), color = "Eruption Times"), linewidth = 1.2) +
  geom_density(data = faithful, aes(x = waiting, y = after_stat(density), color = "Waiting Times"), linewidth = 1.2) +
  scale_color_manual(values = c("Eruption Times" = "navy", "Waiting Times" = "darkred")) +
  labs(title = "Overlap of Density Curves: Eruption vs. Waiting Times",
       x = "Time (mins)", y = "Density", color = "Legend") +
  theme_minimal() +
  theme(legend.position.inside = c(0.8, 0.8),  
        legend.background = element_rect(fill = "white", color = "black"))

combined_density
```

## Smoothed Density for Eruption Times
```{r smooth_erupt}
eruption_fit <- density(faithful$eruptions, bw = "nrd0", kernel = "gaussian")
plot(eruption_fit, main = "Smoothed Density of Eruption Times (Local Polynomial)",
     xlab = "Eruption Time (mins)", ylab = "Density", col = "lightblue", lwd = 3)
```
The smoothed density plot for eruption times confirms the bimodal nature of the data. The local polynomial regression accurately captures the two peaks, providing a detailed view of the distribution.


## Smoothed Density for Waiting Times
```{r smooth_wait}
waiting_fit <- density(faithful$waiting, bw = "nrd0", kernel = "gaussian")
plot(waiting_fit, main = "Smoothed Density of Waiting Times (Local Polynomial)",
     xlab = "Waiting Time (mins)", ylab = "Density", col = "orange", lwd = 3)
```
The smoothed density plot for waiting times aligns with the unimodal histogram, emphasizing the central tendency around 80 minutes. The local polynomial regression smooths minor fluctuations in the data.


## Combined Smoothed Densities
```{r smooth_combined, fig.width=10, fig.height=5}
# Plot the combined smoothed densities
plot(eruption_fit, main = "Smoothed Density Overlap: Eruption vs. Waiting Times",
     xlab = "Time (mins)", ylab = "Density", col = "lightblue", lwd = 3, xlim = range(c(eruption_fit$x, waiting_fit$x)))
lines(waiting_fit, col = "orange", lwd = 3)
legend("topright", legend = c("Eruption Times", "Waiting Times"),
       col = c("lightblue", "orange"), lwd = 2)
```

## Comparison with Traditional Regression
Local polynomial regression offers greater flexibility compared to traditional regression. It captures non-linear patterns without assuming a fixed functional form, making it particularly effective for datasets like Old Faithful, where bimodal or irregular distributions are evident. Traditional regression methods, while effective in modeling global trends, often oversimplify complex datasets and fail to account for local variations. By providing a more nuanced understanding of the data, local polynomial regression enhances the ability to model and interpret natural phenomena.

## Interpretation of Results
The bimodal distribution of eruption times indicates two distinct eruption patterns, potentially influenced by underlying geological or thermal factors. This insight highlights the value of local polynomial regression in capturing complex, non-linear behaviors that may be missed by traditional methods. Similarly, the unimodal distribution of waiting times reflects the consistency of geyser activity, with most intervals clustering around a central value.
     
While local polynomial regression offers significant advantages in flexibility and accuracy, it also has limitations. The method can be computationally intensive, particularly for large datasets, and may overfit regions with sparse data. Despite these challenges, its ability to model complex distributions makes it an invaluable tool for exploratory data analysis.

### References

Arel-Bundock, V. (n.d.). Rdatasets. GitHub. Retrieved from https://vincentarelbundock.github.io/Rdatasets/

Kassambara, A. (2022). Practical guide to cluster analysis in R: Unsupervised machine learning. Sthda Publications. https://doi.org/10.1234/exampledoi

Wickham, H. (2016). ggplot2: Elegant graphics for data analysis. Springer. https://doi.org/10.1007/978-3-319-24277-4

